Vagueness \& Rationality

### 2018-10-27 manuscript/v-book-2018.tex
replaced double quotes with single quotes throughout

\chapter{The Elusive Benefits of Vagueness: Evidence from Experiments}
Much of everyday language is vague, even in situations where vagueness could have been avoided (i.e., where vagueness is used ``strategically'').
Yet the benefits of vagueness for hearers and readers are proving to be elusive.
We discuss a range of earlier controlled experiments with human participants, and we report on a new series of experiments that we ourselves have conducted in recent years.
These experiments, which focus on vague expressions that are part of referential noun phrases, aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness.
After presenting the evidence, we argue that it supports a view where the benefits that vague terms exert are due to other influences, and not to vagueness itself.

\section{Introduction}
In academic use, the word `vagueness' has a specific meaning.
Keefe and Smith, for example, state that ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \citep[p.~4]{keefe1997vagueness}, as do \citet{EgreKlinedinst}.
The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \citet[p.~1]{lipmanvague}.
A typical example is the word ``tall'', as applied to people for example, because there is no precise, known height which separates those who are tall from those who are not.
The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

Linguists, philosophers of language, and more recently game theorists have asked why natural languages contain so many vague expressions \citet{Lipman:2000fk, lipmanvague}, which are
often used even in situations where the speaker could have used an expression that is not vague (i.e., crisp); in these situations we say that vagueness is used \emph{strategically}.
By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \citet[][p.~1]{lipmanvague}.
Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport.
In such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall'').
Lipman uses this scenario to explain why standard game theory models of communication \citep[e.g.,][]{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs.

Lipman argues that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there were advantages as well as disadvantages associated with vague expressions.
Lipman asks, essentially, what these advantages might be.
Several tentative answers to Lipman's question have been offered \citep[see][]{van2009utility, vanDeemterBook}.
One of the most promising answers appears to be the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \citep[e.g.,][]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}.
For example, \citeauthor{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze'' (\citeyear[][p.~11]{lipmanvague}).
We shall refer to this as the \emph{cost reduction} hypothesis.

This article brings an experimental approach to these issues, focussing on vagueness in descriptions (e.g., ``the square with few dots'') and its effect on the hearer's ability to act on a given description, as measured by the time that it takes hearers to click on the referent of a description.\footnote{Other metrics could have been chosen, such as hearers' ability to remember information, for example, or error rates. Although error rates play a minor role in the present paper, for reasons that will become clear, we focus on response times in particular.}
We find that, although hearer benefits from vague descriptions are straightforward to demonstrate in many cases, a closer experimental analysis militates against the conclusion that vagueness itself --- as defined above, in terms of the existence of borderline cases --- lies at the heart of these results.
Instead, it is other factors, such as the presence of an overt numerical expression in the description, that proved to be decisive.
We believe that, despite the fact that our experiments are unavoidably focussed on a specific class of vague expressions (since any experiment can only deal with a limited number of different stimuli), these findings are potentially important, because they call into question whether ``strategic'' vagueness (i.e., vagueness where the speaker had a choice, because she could have produced a crisp expression instead) has any advantages at all.
In other words, returning to Lipman's question, it is possible that vagueness has evolved partly as a necessary evil (e.g., because of the limits of observation and prediction) and partly as a side effect of other factors.

\section{Related work}\label{related-work}
A number of answers to Lipman's question have been proposed, but as we have argued elsewhere, it appears that some of these do not to hold up to scrutiny \citep{van2009utility,vanDeemterBook}. For example,
\begin{itemize}
\item It has been proposed that the existence of vague words makes a language more efficient \citep{barwiseperry}. The idea is that vagueness allows the use of one and the same word (e.g., ``big'') in different situations, which require different standards. For instance, the vague word ``big'' can denote a mouse as well as an elephant. The idea is plausible, but the problem is that ``big'' is efficient not because it is vague (i.e., not because it allows borderline cases) but because it is context dependent. To see this, note for example that superlatives are not vague (i.e., they do not permit borderline cases), but they {\em are} context dependent: ``the biggest $x$'' ascribes different sizes to the referent depending on $x$. It is for this reason that we can use the word ``biggest''  to talk about the biggest mouse, the biggest elephant, and so on. Context-dependence, however, does not imply vagueness.
\item Vague words are capable of combining a statement of quantity with an evaluative statement \citep{veltman}. The idea is that when we say ``The patient has a fever,'' we do not merely assert that her temperature lies above a certain value, we also imply that the deviation is clinically significant (i.e., something is not right with the patient). The problem with this plausible idea is that evaluation does not imply vagueness. For example, the medical term ``obese'' is evaluative (i.e., it is not healthy to be obese), yet its standard medical definition in terms of Body Mass Index is perfectly crisp (i.e., without borderline cases), with all and only BMI values above 30 counting as obese.
\end{itemize}
Analogous observations can be made about a number of other purported benefits of vagueness; one of the few hypotheses left standing at the moment is the {\em cost reduction} hypothesis, which we mentioned in the previous section. The findings on which we are reporting in the present paper will follow a familiar pattern: we will explore the cost reduction hypothesis, only to conclude that it may not stand up to scrutiny. Our approach, this time, will be experimental: we conduct controlled experiments to investigate the effect that vague expressions have on a hearer.

@@

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range --- the end closest to the ideal HHI scores. This ``illusion of proximity'' \citep[][p.~4]{Mishra01042011} to the goal is argued to have allowed participants to generate positive expectancies that lead to behaviours that improved performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \citep{keefe1997vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeauthor{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeauthor{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

@@

It is, however, questionable whether the ``evaluative categories'' manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60--70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.

@@

In Experiment 1, reported below, we also addressed another issue with the second pilot experiment: participants chose one of two dot arrays --- therefore the `vague' quantifiers (few and many) uniquely identified one square. Recall our definition of vague --- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''. The quantifiers `few' and `many' might not have realised their potential for vagueness in the context of a choice between (only) two alternatives --- where there is no borderline alternative that could result in few and many being ``not well-defined''. Furthermore, using definite articles in the instructions may have reinforced the participant's impression that only one choice counted as correct, and this impression could have been reinforced by the use of error feedback.

@@

\begin{itemize}
	\item In the vague numerical condition the instruction was ``Choose a square with about 10 dots'' --- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response.
	\item In the vague verbal condition we used ``Choose a square with few dot''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item In the crisp numerical condition we used ``Choose the square with 6 dots''. The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item For crisp verbal, we used ``Choose the square with the fewest dots''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., ``Choose a square with about 10 dots''). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as ``correct'' or ``incorrect'', but instead as ``expected response''; ``borderline response''; and ``extreme response'', and we did not want to draw participants' attention to this distinction explicitly. The participant's choice was recorded for analysis.

\begin{table}[htbp]
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots''}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

\section{Experiment 2: Focus on instructions that contain numerals}
The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., whether the task amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}). We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table~\ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded. Table~\ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

\begin{table}
\centering
\caption{Experiment 2: Instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots''}
\label{Instructions for e2}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Selection & Crisp & Vague \\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Comparison & fewer than 20 dots & far fewer than 20 dots \\
06:15:24 & Small & Matching & 6 dots & about 10 dots \\
06:15:24 & Large & Comparison & more than 10 dots & far more than 10 dots \\
06:15:24 & Large & Matching & 24 dots & about 20 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Comparison & fewer than 30 dots & far fewer than 30 dots \\
16:25:34 & Small & Matching & 16 dots & about 20 dots \\
16:25:34 & Large & Comparison & more than 20 dots & far more than 20 dots \\
16:25:34 & Large & Matching & 34 dots & about 30 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Comparison & fewer than 40 dots & far fewer than 40 dots \\
26:35:44 & Small & Matching & 26 dots & about 30 dots \\
26:35:44 & Large & Comparison & more than 30 dots & far more than 30 dots \\
26:35:44 & Large & Matching & 44 dots & about 40 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Comparison & fewer than 50 dots & far fewer than 50 dots \\
36:45:54 & Small & Matching & 36 dots & about 40 dots \\
36:45:54 & Large & Comparison & more than 40 dots & far more than 40 dots \\
36:45:54 & Large & Matching & 54 dots & about 50 dots \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus, showing the sequence of presentation, with the instruction screen at the top, followed by the presentation of the ``target'' array, followed by the stimulus to which participants responded.}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions for the smallest array, arranged by condition. The instructions given in the table started with ``Choose a square with \ldots''}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

@@

The benefits of vagueness in the \emph{comparison} task in Experiments 3 and 4 could thus be explained as differences in the number of valid targets for the expression. This leads us to speculate that the benefit for vagueness here could be due to the vague expression foregrounding a particular valid target, while the crisp expression carries with it the additional task of distinguishing between two alternative valid targets, something we propose to call a ``range-reduction'' benefit.

@@

First, the experiments by \citeauthor{Mishra01042011} suggest that it is important how much information is conveyed by an expression, and their findings are echoed by our own thoughts about range reduction (following Experiments 2 and 3). In the case of (a)--(d) above, it appears that (a) conveys the most detailed information (designating the smallest segment of the temperature scale), followed by (b), then (d), then (c) (e.g., 40 degrees is above 25, but at 40 Celsius the word ``warm'' is likely to give way to ``hot'' or ``scorching''):
%
\begin{quote}
$a < b < d < c$
\end{quote}
%
If these hunches are correct, then it seems to us that it is relatively unimportant whether a given expression is vague or crisp. Other factors seem more important; moreover, it may depend on the task and the audience which of a--d is preferred. For example, an \emph{expert} may prefer to read expression (a), because it gives her the most detailed information on which to base her decisions. On the other hand, expression (d) (``warm'') is shorter than the other three and avoids the use of numbers; our experiments suggest that this may make (d) more rapidly understood than its competitors; earlier experiments point in the same direction, given the evaluative nature of ``warm'' (recall section~\ref{related-work}), which is especially important if the hearer is unfamiliar with the metric used. These considerations suggest that {\em non-experts} might prefer expression (d).

>>>>

\chapter{The Elusive Benefits of Vagueness: Evidence from Experiments}
Much of everyday language is vague, even in situations where vagueness could have been avoided (i.e., where vagueness is used `strategically').
Yet the benefits of vagueness for hearers and readers are proving to be elusive.
We discuss a range of earlier controlled experiments with human participants, and we report on a new series of experiments that we ourselves have conducted in recent years.
These experiments, which focus on vague expressions that are part of referential noun phrases, aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness.
After presenting the evidence, we argue that it supports a view where the benefits that vague terms exert are due to other influences, and not to vagueness itself.

\section{Introduction}
In academic use, the word `vagueness' has a specific meaning.
Keefe and Smith, for example, state that `vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes' \citep[p.~4]{keefe1997vagueness}, as do \citet{EgreKlinedinst}.
The crucial criterion is the existence of borderline cases: `a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise' \citet[p.~1]{lipmanvague}.
A typical example is the word `tall', as applied to people for example, because there is no precise, known height which separates those who are tall from those who are not.
The crucial point is that `tall' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

Linguists, philosophers of language, and more recently game theorists have asked why natural languages contain so many vague expressions \citet{Lipman:2000fk, lipmanvague}, which are
often used even in situations where the speaker could have used an expression that is not vague (i.e., crisp); in these situations we say that vagueness is used \emph{strategically}.
By introducing borderline cases, these expressions create potential misunderstandings, thereby creating `a worldwide several-thousand year efficiency loss' \citet[][p.~1]{lipmanvague}.
Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport.
In such a scenario, a precise description of the person's height (e.g., `The person's height is 187.96 cm') would be more useful than a vague one (`The person is tall').
Lipman uses this scenario to explain why standard game theory models of communication \citep[e.g.,][]{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs.

Lipman argues that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there were advantages as well as disadvantages associated with vague expressions.
Lipman asks, essentially, what these advantages might be.
Several tentative answers to Lipman's question have been offered \citep[see][]{van2009utility, vanDeemterBook}.
One of the most promising answers appears to be the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \citep[e.g.,][]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}.
For example, \citeauthor{lipmanvague} writes: `For the listener, information which is too specific may require more effort to analyze' (\citeyear[][p.~11]{lipmanvague}).
We shall refer to this as the \emph{cost reduction} hypothesis.

This article brings an experimental approach to these issues, focussing on vagueness in descriptions (e.g., `the square with few dots') and its effect on the hearer's ability to act on a given description, as measured by the time that it takes hearers to click on the referent of a description.\footnote{Other metrics could have been chosen, such as hearers' ability to remember information, for example, or error rates. Although error rates play a minor role in the present paper, for reasons that will become clear, we focus on response times in particular.}
We find that, although hearer benefits from vague descriptions are straightforward to demonstrate in many cases, a closer experimental analysis militates against the conclusion that vagueness itself --- as defined above, in terms of the existence of borderline cases --- lies at the heart of these results.
Instead, it is other factors, such as the presence of an overt numerical expression in the description, that proved to be decisive.
We believe that, despite the fact that our experiments are unavoidably focussed on a specific class of vague expressions (since any experiment can only deal with a limited number of different stimuli), these findings are potentially important, because they call into question whether `strategic' vagueness (i.e., vagueness where the speaker had a choice, because she could have produced a crisp expression instead) has any advantages at all.
In other words, returning to Lipman's question, it is possible that vagueness has evolved partly as a necessary evil (e.g., because of the limits of observation and prediction) and partly as a side effect of other factors.

\section{Related work}\label{related-work}
A number of answers to Lipman's question have been proposed, but as we have argued elsewhere, it appears that some of these do not to hold up to scrutiny \citep{van2009utility,vanDeemterBook}. For example,
\begin{itemize}
\item It has been proposed that the existence of vague words makes a language more efficient \citep{barwiseperry}. The idea is that vagueness allows the use of one and the same word (e.g., `big') in different situations, which require different standards. For instance, the vague word `big' can denote a mouse as well as an elephant. The idea is plausible, but the problem is that `big' is efficient not because it is vague (i.e., not because it allows borderline cases) but because it is context dependent. To see this, note for example that superlatives are not vague (i.e., they do not permit borderline cases), but they {\em are} context dependent: `the biggest $x$' ascribes different sizes to the referent depending on $x$. It is for this reason that we can use the word `biggest'  to talk about the biggest mouse, the biggest elephant, and so on. Context-dependence, however, does not imply vagueness.
\item Vague words are capable of combining a statement of quantity with an evaluative statement \citep{veltman}. The idea is that when we say `The patient has a fever,' we do not merely assert that her temperature lies above a certain value, we also imply that the deviation is clinically significant (i.e., something is not right with the patient). The problem with this plausible idea is that evaluation does not imply vagueness. For example, the medical term `obese' is evaluative (i.e., it is not healthy to be obese), yet its standard medical definition in terms of Body Mass Index is perfectly crisp (i.e., without borderline cases), with all and only BMI values above 30 counting as obese.
\end{itemize}
Analogous observations can be made about a number of other purported benefits of vagueness; one of the few hypotheses left standing at the moment is the {\em cost reduction} hypothesis, which we mentioned in the previous section. The findings on which we are reporting in the present paper will follow a familiar pattern: we will explore the cost reduction hypothesis, only to conclude that it may not stand up to scrutiny. Our approach, this time, will be experimental: we conduct controlled experiments to investigate the effect that vague expressions have on a hearer.

@@

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range --- the end closest to the ideal HHI scores. This `illusion of proximity' \citep[][p.~4]{Mishra01042011} to the goal is argued to have allowed participants to generate positive expectancies that lead to behaviours that improved performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \citep{keefe1997vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeauthor{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeauthor{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

@@

It is, however, questionable whether the `evaluative categories' manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60--70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.

@@

In Experiment 1, reported below, we also addressed another issue with the second pilot experiment: participants chose one of two dot arrays --- therefore the `vague' quantifiers (few and many) uniquely identified one square. Recall our definition of vague --- `a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'. The quantifiers `few' and `many' might not have realised their potential for vagueness in the context of a choice between (only) two alternatives --- where there is no borderline alternative that could result in few and many being `not well-defined'. Furthermore, using definite articles in the instructions may have reinforced the participant's impression that only one choice counted as correct, and this impression could have been reinforced by the use of error feedback.

@@

\begin{itemize}
	\item In the vague numerical condition the instruction was `Choose a square with about 10 dots' --- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response.
	\item In the vague verbal condition we used `Choose a square with few dot'. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item In the crisp numerical condition we used `Choose the square with 6 dots'. The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item For crisp verbal, we used `Choose the square with the fewest dots'. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., `Choose a square with about 10 dots'). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as `correct' or `incorrect', but instead as `expected response'; `borderline response'; and `extreme response', and we did not want to draw participants' attention to this distinction explicitly. The participant's choice was recorded for analysis.

\begin{table}[htbp]
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with `Choose \ldots'}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

\section{Experiment 2: Focus on instructions that contain numerals}
The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., whether the task amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure~\ref{Experiment1and2examplestimulus}). We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table~\ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded. Table~\ref{Instructions for e2} shows the instructions for each condition. Note the difference between `fewer than 20' and `far fewer than 20': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

\begin{table}
\centering
\caption{Experiment 2: Instructions arranged by condition. The instructions given in the table started with `Choose a square with \ldots'}
\label{Instructions for e2}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Selection & Crisp & Vague \\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Comparison & fewer than 20 dots & far fewer than 20 dots \\
06:15:24 & Small & Matching & 6 dots & about 10 dots \\
06:15:24 & Large & Comparison & more than 10 dots & far more than 10 dots \\
06:15:24 & Large & Matching & 24 dots & about 20 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Comparison & fewer than 30 dots & far fewer than 30 dots \\
16:25:34 & Small & Matching & 16 dots & about 20 dots \\
16:25:34 & Large & Comparison & more than 20 dots & far more than 20 dots \\
16:25:34 & Large & Matching & 34 dots & about 30 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Comparison & fewer than 40 dots & far fewer than 40 dots \\
26:35:44 & Small & Matching & 26 dots & about 30 dots \\
26:35:44 & Large & Comparison & more than 30 dots & far more than 30 dots \\
26:35:44 & Large & Matching & 44 dots & about 40 dots \\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Comparison & fewer than 50 dots & far fewer than 50 dots \\
36:45:54 & Small & Matching & 36 dots & about 40 dots \\
36:45:54 & Large & Comparison & more than 40 dots & far more than 40 dots \\
36:45:54 & Large & Matching & 54 dots & about 50 dots \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus, showing the sequence of presentation, with the instruction screen at the top, followed by the presentation of the `target' array, followed by the stimulus to which participants responded.}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions for the smallest array, arranged by condition. The instructions given in the table started with `Choose a square with \ldots'}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

@@

The benefits of vagueness in the \emph{comparison} task in Experiments 3 and 4 could thus be explained as differences in the number of valid targets for the expression. This leads us to speculate that the benefit for vagueness here could be due to the vague expression foregrounding a particular valid target, while the crisp expression carries with it the additional task of distinguishing between two alternative valid targets, something we propose to call a `range-reduction' benefit.

@@

First, the experiments by \citeauthor{Mishra01042011} suggest that it is important how much information is conveyed by an expression, and their findings are echoed by our own thoughts about range reduction (following Experiments 2 and 3). In the case of (a)--(d) above, it appears that (a) conveys the most detailed information (designating the smallest segment of the temperature scale), followed by (b), then (d), then (c) (e.g., 40 degrees is above 25, but at 40 Celsius the word `warm' is likely to give way to `hot' or `scorching'):
%
\begin{quote}
$a < b < d < c$
\end{quote}
%
If these hunches are correct, then it seems to us that it is relatively unimportant whether a given expression is vague or crisp. Other factors seem more important; moreover, it may depend on the task and the audience which of a--d is preferred. For example, an \emph{expert} may prefer to read expression (a), because it gives her the most detailed information on which to base her decisions. On the other hand, expression (d) (`warm') is shorter than the other three and avoids the use of numbers; our experiments suggest that this may make (d) more rapidly understood than its competitors; earlier experiments point in the same direction, given the evaluative nature of `warm' (recall section~\ref{related-work}), which is especially important if the hearer is unfamiliar with the metric used. These considerations suggest that {\em non-experts} might prefer expression (d).

One way to see why vagueness (as defined in our Introduction) may not benefit human communication is the following thought experiment. Suppose a group of speakers understand the word `warm' as vague, agreeing that temperatures above 26 count as warm, and temperatures below 24 do not count as warm, but considering temperatures between 24 and 26 as borderline cases. Now one day these speakers agree to sharpen up their definition, deciding that, henceforth, `warm' means `$>25$ degrees' (as in (c) above): this decision resolves the borderline cases, while everything else remains the same. It seems unlikely that this change in language use, from a vague meaning to a crisp one (i.e., one that has no borderline cases anymore), would lower the utility of the word. Our experimental findings, and the conclusions that we draw from them, are consistent with this idea.
